---
title: "Report on Elastic net based PTRS"
---

```{r setup}
library(dplyr)
library(data.table)
library(pander)
library(ggplot2)
panderOptions('table.split.table', Inf)
theme_set(theme_bw(base_size=12))
source('https://gist.githubusercontent.com/liangyy/43912b3ecab5d10c89f9d4b2669871c9/raw/8151c6fe70e3d4ee43d9ce340ecc0eb65172e616/my_ggplot_theme.R')
th$panel.border = element_rect(colour = th$axis.line$colour)
blood_traits = c('wbc', 'rbc', 'platelet', 'lymphocyte', 'monocyte', 'neutrophil', 'eosinophil', 'basophil')
color_mixer = c('African' = '#E74C3C', 'British_test' = '#28B463', 'British_validation' = '#D4AC0D', 'Chinese' = '#3498DB', 'Indian' = '#9B59B6', 'British_insample' = '#AAB7B8')
my_parser = function(x) {
  as.numeric(stringr::str_remove(x, 'pval_'))
}
```

# Overview

Here we implemented elastic net based PTRS predictor along regularization path. 
We are interested in the performance of EN PTRS as compared to PTRS based on least squares and PTRS based on S-PrediXcan.

# Load data

```{r data}
df_ls = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-least_square_model-ukb_imp_x_ctimp_Whole_Blood_British.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_en0 = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-elastic_net_British.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_en2 = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-elastic_net_British_finer.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_en = rbind(df_en0, df_en2) %>% arrange(desc(lambda))
df_spredixcan = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-spredixcan_ptrs.subset1_British.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_prs = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-prs.subset1_British.tsv', header = T, sep = '\t', stringsAsFactors = F)

df_spredixcan$sample[df_spredixcan$sample == 'British-test-1'] = 'British_test'
df_spredixcan$sample[df_spredixcan$sample == 'British-validation-1'] = 'British_validation'
df_prs$sample[df_prs$sample == 'British-test-1'] = 'British_test'
df_prs$sample[df_prs$sample == 'British-validation-1'] = 'British_validation'
df_spredixcan = df_spredixcan %>% mutate(cutoff = my_parser(ptrs_cutoff))
df_prs = df_prs %>% rename(prs_cutoff = ptrs_cutoff)
df_prs = df_prs %>% arrange(desc(-prs_cutoff))
```

# EN PTRS along regularization path

```{r en, fig.width=5, fig.height=10}
df_en %>% filter(trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + geom_hline(data = df_ls %>% filter(trait %in% blood_traits, sample != 'British_insample'), aes(yintercept = partial_r2, color = sample)) + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS Blood traits')
df_en %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + geom_hline(data = df_ls %>% filter(!trait %in% blood_traits, sample != 'British_insample'), aes(yintercept = partial_r2, color = sample)) + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS Non-blood traits')
```

**Take-away**: 
Except for very occasional cases, regularization dose not help increase the transferability.
It is possible that our current lambda grid is too crude.
**TODO**: I will use a finer grid.

# S-PrediXcan PTRS 

```{r spredixcan, fig.width=5, fig.height=10}
df_spredixcan %>% filter(trait %in% blood_traits) %>% ggplot() + geom_path(aes(x = log(cutoff), y = partial_r2, color = sample)) + facet_grid(trait~ptrs_type, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('S-PrediXcan PTRS Blood traits')
df_spredixcan %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(cutoff), y = partial_r2, color = sample)) + facet_grid(trait~ptrs_type, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('S-PrediXcan PTRS Non-blood traits')
```

**Take-away**:
No clear pattern to recognize. 
Generally speaking, the best cutoff could depend on trait.

# LD-clump PRS

```{r prs, fig.width=6, fig.height=4}
df_prs %>% filter(trait %in% blood_traits) %>% ggplot() + geom_path(aes(x = log(prs_cutoff), y = partial_r2, color = sample)) + facet_wrap(~trait, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('PRS Blood traits')
df_prs %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(prs_cutoff), y = partial_r2, color = sample)) + facet_wrap(~trait, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('PRS Non-blood traits')
```

**Take-away**:
Best British cutoff is less stringent than best cutoff for other populations.

# Best EN PTRS, S-PrediXcan PTRS, and LD-clump PRS

```{r best}
best_en = rbind(
  df_ls %>% select(sample, trait, partial_r2),
  df_en %>% select(sample, trait, partial_r2)
) %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
best_spredixcan = df_spredixcan %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
best_prs = df_prs %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
```

## EN PTRS vs. S-PrediXcan PTRS

```{r best_vs1, fig.width=6, fig.height=4}
merge = inner_join(best_en, best_spredixcan, by = c('sample', 'trait'), suffix = c('.en', '.spredixcan'))
merge %>% ggplot() + geom_point(aes(x = r2_max.spredixcan, y = r2_max.en)) + facet_wrap(~sample, scales = 'free') + geom_abline(intercept = 0, slope = 1) + th
```

**Take-away**:
This comparison indicates the need to build multivariate prediction model.

## EN PTRS vs. PRS

```{r best_vs2, fig.width=6, fig.height=4}
merge = inner_join(best_en, best_prs, by = c('sample', 'trait'), suffix = c('.en', '.prs'))
merge %>% ggplot() + geom_point(aes(x = r2_max.prs, y = r2_max.en)) + facet_wrap(~sample, scales = 'free') + geom_abline(intercept = 0, slope = 1) + th
```

**Take-away**:
EN PTRS has comparable scale to PRS.

# Heritability vs PVS in British

```{r h2_pve, fig.width=4, fig.height=4}
h2 = readRDS('~/Documents/repo/github/ptrs-ukb/analysis_output/hsq_neale_lab.rds')
pve = readRDS('~/Documents/repo/github/ptrs-ukb/analysis_output/regulability_ctimp.rds')
pve$population[pve$population == 'British-test-1'] = 'British_test'
pve$population[pve$population == 'British-validation-1'] = 'British_validation'
merge = inner_join(h2 %>% select(trait, h2_observed, h2_observed_se), pve %>% filter(population == 'British_test'), by = "trait")
delta_mtd = function(mx, vx, my, vy) {
  m = mx / my
  v = vx / (my ^ 2) + vy * (mx ^ 2) / (my ^ 4)
  return(list(m = m, v = v))
}
meta_fixed = function(m, se) {
  w = 1 / (se^2)
  mfe = sum(m * w) / sum(w)
  vfe = 1 / sum(w)
  return(list(m = mfe, se = sqrt(vfe)))
}
ratio = delta_mtd(merge$h_sq, merge$h_sq_se^2, merge$h2_observed, merge$h2_observed_se^2)
merge = merge %>% mutate(ratio_mean = ratio$m, ratio_se = sqrt(ratio$v))
ratio_fe = meta_fixed(merge$ratio_mean, merge$ratio_se)
as.data.frame(ratio_fe, col.names = c('ratio_mean', 'ratio_se')) %>% pander
merge %>% ggplot() + 
  geom_point(aes(x = h2_observed, y = h_sq)) + 
  geom_errorbar(aes(x = h2_observed, ymin = h_sq - 1.96 * h_sq_se, ymax = h_sq + 1.96 * h_sq_se)) + 
  geom_errorbarh(aes(xmin = h2_observed - 1.96 * h2_observed_se, xmax = h2_observed + 1.96 * h2_observed_se, y = h_sq)) +
  th +
  geom_abline(intercept = 0, slope = ratio_fe$m + 1.96 * ratio_fe$se) + 
  geom_abline(intercept = 0, slope = ratio_fe$m - 1.96 * ratio_fe$se)
```

**Take-away**: 
PVE is about 20% of heritability.

# PVE vs PTRS

```{r pve_ptrs, fig.width=4, fig.height=4}
merge = inner_join(pve, rbind(best_en %>% mutate(type = 'EN'), best_spredixcan %>% mutate(type = 'S-PrediXcan')), by = c('trait', 'population' = 'sample')) 
merge %>% ggplot() + geom_point(aes(x = h_sq, y = r2_max, color = type)) + facet_wrap(~population) + th + ggtitle('PVE vs PTRS') + geom_abline(slope = 1, intercept = 0) + theme(legend.position = 'bottom')
```

**Take-away**: 
EN PTRS is better predictor than S-PrediXcan PTRS in general.

# Heritability vs R2 (British only)

```{r h2_prs, fig.width=4, fig.height=4}
merge = inner_join(h2, rbind(best_en %>% mutate(type = 'EN PTRS'), best_spredixcan %>% mutate(type = 'S-PrediXcan PTRS'), best_prs %>% mutate(type = 'PRS')) %>% filter(sample == 'British_test'), by = c('trait')) 
merge %>% ggplot() + geom_point(aes(x = h2_observed, y = r2_max, color = type)) + th + ggtitle('PTRS/PRS vs heritability') + geom_abline(slope = 1, intercept = 0) + theme(legend.position = 'bottom')
```

**Take-away**: 
LD-clump based PRS is not optimal. 

# Transferability

```{r trans}
trans_en = best_en %>% left_join(best_en %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
trans_spredixcan = best_spredixcan %>% left_join(best_spredixcan %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
trans_prs = best_prs %>% left_join(best_prs %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
rbind(
  trans_en %>% mutate(type = 'EN PTRS'),
  trans_spredixcan %>% mutate(type = 'S-PrediXcan PTRS'), 
  trans_prs %>% mutate(type = 'PRS')
) %>% filter(sample != 'British_insample') %>%
  ggplot() + geom_violin(aes(x = sample, y = transferability, color = type), position = position_dodge(width = 0.5)) + geom_boxplot(aes(x = sample, y = transferability, color = type), position = position_dodge(width = 0.5), width = 0.1) + th + theme(legend.position = 'bottom')
  
```

**Take-away**: 
By using EN, African, Chinese, and Indian do not benefit (as comparing to S-PrediXcan approach) more than British. So that EN PTRS has lower transferability than S-PrediXcan PTRS. 
