---
title: "Report on Elastic net based PTRS (MESA)"
---

```{r setup}
library(dplyr)
library(data.table)
library(pander)
library(ggplot2)
panderOptions('table.split.table', Inf)
theme_set(theme_bw(base_size=12))
source('https://gist.githubusercontent.com/liangyy/43912b3ecab5d10c89f9d4b2669871c9/raw/8151c6fe70e3d4ee43d9ce340ecc0eb65172e616/my_ggplot_theme.R')
th$panel.border = element_rect(colour = th$axis.line$colour)
blood_traits = c('wbc', 'rbc', 'platelet', 'lymphocyte', 'monocyte', 'neutrophil', 'eosinophil', 'basophil')
color_mixer = c('African' = '#E74C3C', 'British_test' = '#28B463', 'British_validation' = '#D4AC0D', 'Chinese' = '#3498DB', 'Indian' = '#9B59B6', 'British_insample' = '#AAB7B8')
my_parser = function(x) {
  as.numeric(stringr::str_remove(x, 'pval_'))
}
source('https://raw.githubusercontent.com/liangyy/ptrs-ukb/master/code/rlib_doc.R?token=AC7RPMKFRGMEYNIAT5QTDTK6IRTOM')
```

# Overview

We trained PTRS models using the genes occurring in both MESA CAU and MESA AFHI. 
The EN PTRS models were trained using EUR individuals with MESA CAU. 
The EN PTRS in all populations was calculated using MESA CAU and MESA AFHI respectively.

# Load data

```{r data}
df_cau = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-elastic_net_MESA_CAU_British_in_CAU.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_afhi = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-elastic_net_MESA_CAU_British_in_AFHI.tsv', header = T, sep = '\t', stringsAsFactors = F)
df_en = rbind(df_cau %>% mutate(pred_expr = 'CAU'), df_afhi %>% mutate(pred_expr = 'AFHI')) %>% arrange(desc(lambda))
df_prs = read.table('~/Desktop/tmp/ptrs-tf/from_nucleus/partial_r2-prs.subset1_British.tsv', header = T, sep = '\t', stringsAsFactors = F)

df_prs$sample[df_prs$sample == 'British-test-1'] = 'British_test'
df_prs$sample[df_prs$sample == 'British-validation-1'] = 'British_validation'
df_prs = df_prs %>% rename(prs_cutoff = ptrs_cutoff)
df_prs = df_prs %>% arrange(desc(-prs_cutoff))
```

# EN PTRS along regularization path

```{r en, fig.width=5, fig.height=10}
df_cau %>% filter(trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS (CAU) Blood traits')
df_cau %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS (CAU) Non-blood traits')
```

```{r en2, fig.width=5, fig.height=10}
df_afhi %>% filter(trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS (CAU) Blood traits')
df_afhi %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(lambda), y = partial_r2, color = sample)) + facet_grid(trait~alpha, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('EN PTRS (CAU) Non-blood traits')
```

# LD-clump PRS

```{r prs, fig.width=6, fig.height=4}
df_prs %>% filter(trait %in% blood_traits) %>% ggplot() + geom_path(aes(x = log(prs_cutoff), y = partial_r2, color = sample)) + facet_wrap(~trait, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('PRS Blood traits')
df_prs %>% filter(!trait %in% blood_traits, sample != 'British_insample') %>% ggplot() + geom_path(aes(x = log(prs_cutoff), y = partial_r2, color = sample)) + facet_wrap(~trait, scales = 'free_y') + th + 
  theme(legend.position = 'bottom') + 
  scale_color_manual(values = color_mixer) + 
  ggtitle('PRS Non-blood traits')
```


# Best EN PTRS, and LD-clump PRS

```{r best}
best_en = df_en %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
best_cau = df_cau %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
best_afhi = df_afhi %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
best_prs = df_prs %>% group_by(trait, sample) %>% summarize(r2_max = max(partial_r2, na.rm = T)) %>% ungroup()
```

## EN PTRS CAU vs AFHI

```{r best_vs3, fig.width=6, fig.height=4}
merge = inner_join(best_cau, best_afhi, by = c('sample', 'trait'), suffix = c('.cau', '.afhi'))
merge %>% ggplot() + geom_point(aes(x = r2_max.cau, y = r2_max.afhi)) + facet_wrap(~sample, scales = 'free') + geom_abline(intercept = 0, slope = 1) + th
```

## EN PTRS vs. PRS

```{r best_vs2, fig.width=6, fig.height=4}
merge = inner_join(best_en, best_prs, by = c('sample', 'trait'), suffix = c('.en', '.prs'))
merge %>% ggplot() + geom_point(aes(x = r2_max.prs, y = r2_max.en)) + facet_wrap(~sample, scales = 'free') + geom_abline(intercept = 0, slope = 1) + th
```


# Heritability vs PVE in British

```{r h2_pve, fig.width=4, fig.height=4}
h2 = readRDS('~/Documents/repo/github/ptrs-ukb/analysis_output/hsq_neale_lab.rds')
pve = readRDS('~/Documents/repo/github/ptrs-ukb/analysis_output/regulability_mesa_and_ctimp.rds')
pve = pve %>% filter(model == 'CAU')
pve$population[pve$population == 'British'] = 'British_test'
# pve$population[pve$population == 'British-validation-1'] = 'British_validation'
merge = inner_join(h2 %>% select(trait, h2_observed, h2_observed_se), pve %>% filter(population == 'British_test'), by = "trait")
ratio = delta_mtd(merge$h_sq, merge$h_sq_se^2, merge$h2_observed, merge$h2_observed_se^2)
merge = merge %>% mutate(ratio_mean = ratio$m, ratio_se = sqrt(ratio$v))
ratio_fe = meta_fixed(merge$ratio_mean, merge$ratio_se)
as.data.frame(ratio_fe, col.names = c('ratio_mean', 'ratio_se')) %>% pander
merge %>% ggplot() + 
  geom_point(aes(x = h2_observed, y = h_sq)) + 
  geom_errorbar(aes(x = h2_observed, ymin = h_sq - 1.96 * h_sq_se, ymax = h_sq + 1.96 * h_sq_se)) + 
  geom_errorbarh(aes(xmin = h2_observed - 1.96 * h2_observed_se, xmax = h2_observed + 1.96 * h2_observed_se, y = h_sq)) +
  th +
  geom_abline(intercept = 0, slope = ratio_fe$m + 1.96 * ratio_fe$se) + 
  geom_abline(intercept = 0, slope = ratio_fe$m - 1.96 * ratio_fe$se)
```

**Take-away**: 
PVE is about 16% of heritability.

# PVE vs PTRS

```{r pve_ptrs, fig.width=4, fig.height=4}
merge = inner_join(pve, 
                   rbind(
                     best_cau %>% mutate(type = 'CAU'),
                     best_afhi %>% mutate(type = 'AFHI')
                   ), by = c('trait', 'population' = 'sample'))
merge %>% ggplot() + geom_point(aes(x = h_sq, y = r2_max, color = type)) + facet_wrap(~population) + th + ggtitle('PVE vs PTRS') + geom_abline(slope = 1, intercept = 0) + theme(legend.position = 'bottom')
```

# Transferability

```{r trans}
trans_en = best_en %>% left_join(best_en %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
trans_cau = best_cau %>% left_join(best_cau %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
trans_afhi = best_afhi %>% left_join(best_afhi %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
trans_prs = best_prs %>% left_join(best_prs %>% filter(sample == 'British_validation'), by = c('trait'), suffix = c('', '.ref')) %>% mutate(transferability = r2_max / r2_max.ref)
rbind(
  trans_en %>% mutate(type = 'EN PTRS (combined)'),
  trans_cau %>% mutate(type = 'EN PTRS (CAU)'),
  trans_afhi %>% mutate(type = 'EN PTRS (AFHI)'),
  trans_prs %>% mutate(type = 'PRS')
) %>% filter(sample != 'British_insample') %>%
  ggplot() + geom_violin(aes(x = sample, y = transferability, color = type), position = position_dodge(width = 0.5)) + geom_boxplot(aes(x = sample, y = transferability, color = type), position = position_dodge(width = 0.5), width = 0.1) + th + theme(legend.position = 'bottom')
  
```

**Take-away**: 
By using EN, African, Chinese, and Indian do not benefit (as comparing to S-PrediXcan approach) more than British. So that EN PTRS has lower transferability than S-PrediXcan PTRS. 
